Robust Spectroscopic Inference with Imperfect Models

### 2014-11-01 tex/paper.tex
Sys import error on hdf

We have presented an alternative approach to dealing with this fundamental issue, grounded in a 
generative, hierarchical Bayesian framework.  The method advocated here constructs a sophisticated 
likelihood function, employing a non-trivial covariance matrix to treat the correlated 
pixel-to-pixel residuals generated from intrinsically imperfect models.  That matrix is composed of 
a linear combination of {\it global} (stationary) and {\it local} (non-stationary) Gaussian 
process kernels, which parameterize an overall mild covariance structure as well as small patches 
of highly discrepant outlier features.  The approach we describe is generally applicable to any 
spectroscopic inference problem (e.g., population synthesis in unresolved star clusters or 
galaxies, physical/chemical models of emission line spectra in star-forming regions, etc.).  
Moreover, it has the flexibility to incorporate additional information (as priors) or parametric 
complexity (if desired), and could be deployed as a substitute for a simplistic $\chi^2$ metric in 
already-established fitting tools (e.g., {\tt SME}).  To demonstrate how it is used, we determined 
the surface parameters of main-sequence stars with mid-F and mid-M spectral types from high-S/N 
optical and near-infrared data, with reference to pre-computed model libraries 
(Sect.~\ref{sec:examples}).  The source code developed here is open and freely available for use: 
see \url{http://github.com/iancze/StellarSpectra}.

@@

The first step in designing an emulator is to decompose the spectral library into an appropriate basis for emulation \citep{habib07, heitmann09}. We chose the principal component basis (PCA) to decompose the synthetic library into a set of ``eigenspectra,'' following the techniques of \citet{ivezic13}. We decide to truncate our basis to the first $m$ of the eigenspectra, where $m$ is decided by the minimum number of eigenspectra required to reproduce any spectrum in the grid to better than 2\% accuracy for all pixels, however the mean pixel error is generally much smaller ($\lesssim 0.5\%$). An $N_{\rm pix}$ eigenspectrum $\xi_k$, $k = \{1, 2, \ldots, m\}$. Before we do the PCA operation, however, we first standardize the synthetic library by subtracting the mean spectrum and ``whitening'' the spectra by dividing by the standard deviation in each pixel. First, we select only a subset of the spectra synthetic library which will be relevant to our  The mean spectrum is

\begin{figure*}[!t]
  \begin{center}
  \includegraphics{figs/pca_reconstruct.pdf}
  \figcaption{A bunch of reconstrcuted.
\label{fig:pca_reconstruct}}
  \end{center}
\end{figure*}

\begin{equation}
  \xi_\mu = \frac{1}{N} \sum_{i = 1}^N f_\lambda(\{\vt_\ast \}^\textrm{grid})
\end{equation}
and the standard deviation spectrum is
\begin{equation}
  \xi_\sigma = \sqrt{\frac{1}{N} \sum_{i=1}^N \bigl [ f_\lambda(\{\vt_\ast \}^\textrm{grid}) - \xi_\mu \bigr]^2 } 
\end{equation}
We standardize all spectra in the library by subtracting the mean spectrum and dividing by the standard deviation spectrum.
\begin{equation}
  \widetilde{f}_\lambda(\{\vt_\ast \}^\textrm{grid}) = \frac{f_\lambda(\{\vt_\ast \}^\textrm{grid}) - \xi_\mu}{\xi_\sigma}
\end{equation}
Then, the eigenspectra are determined from the standardize grid using PCA analysis (see Ivecik, scikit-learn). To speed computation, we first select $M$ subset spectra from the grid.

[For an example of the eigenspectra for Gl51, see figure A1 (left) and reconstruction of a synthetic spectrum at stellar parameters XX (right).]

Using the PCA basis, we can lossily reconstruct any spectrum from the library by a linear combination of the eigenspectra
\begin{equation}
  f_\lambda(\{\vt_\ast \}^\textrm{grid}) \approx \xi_\mu + \xi_\sigma \sum_{k=1}^m w_k(\{\vt_\ast \}^\textrm{grid}) \xi_k
  \label{eqn:reconstruct}
\end{equation}
where $w_k$ is the weight of a given eigenspectrum, $k$. This weight is a three dimensional scalar function, which depends on the stellar parameters $\vt_\ast$. We can determine the weights at the grid points in the library by taking the dot product of the standardized synthetic spectrum with each of the eigenspectra. For a given eigenspectrum $\xi_k$, the weight is 
\begin{equation}
  w_k(\{\vt_\ast \}^\textrm{grid}) = \sum_\lambda \widetilde{f}_\lambda(\{\vt_\ast \}^\textrm{grid}) \xi_k
\end{equation}
These weights are generally a smooth function of stellar parameters. To simplify the notation, we can write the weights in a length-$m$ column vector
\begin{equation}
  \mathbf{w} = 
  \begin{bmatrix}
    w_1\\
    w_2\\
    \vdots\\
    w_m
  \end{bmatrix}
\end{equation}
and horizontally concatenate the eigenspectra into a matrix with $N_{\rm pix}$ rows and $m$ columns
\begin{equation}
  \mathbf{\Xi} = 
  \begin{bmatrix}
    \xi_1 & \xi_2 & \cdots & \xi_m \\
  \end{bmatrix}
\end{equation}
Therefore we can rewrite Equation~\ref{eqn:reconstruct} as
\begin{equation}
  f_\lambda(\{\vt_\ast \}^\textrm{grid}) \approx \xi_\mu + \xi_\sigma I_{N_{\rm pix}} \mathbf{\Xi} \mathbf{w}
\end{equation}
Where $I_{N_{\rm pix}}$ is an $N_{\rm pix}$ by $N_{\rm pix}$ identity matrix. To summarize, thus far we have decomposed the spectra in the synthetic library into a principal component basis, which allows us to lossily reconstruct any spectrum in the synthetic library as a weighted linear combination of $m$ eigenspectra. The weight corresponding to each eigenspectrum is a relatively-smooth three-dimensional scalar function of the stellar parameters. Therefore, to interpolate a spectrum from the synthetic library for a value of stellar parameters which is not in the spectral library, we must interpolate the value of the weights to an off-grid value. In practice, it might be possible to use a traditional interpolation scheme such as spline interpolation to interpolate the weights directly. However, we found that with high-signal to noise spectra (e.g., Gl51 is S/N $> 400$), the error in any given interpolated representation of the spectrum could constitute a significant portion of the uncertainty budget. Therefore, we decide to develop a Bayesian spectral emulator (cite Habib, Heitmann) which allows us to propagate the uncertainty in interpolation forward to the likelihood calculation. First, we describe how we use a Gaussian process to model the interpolation of the weights then we marginalize over all possible interpolations to arrive at a modified likelihood function.

>>>>

We have presented an alternative approach to dealing with this fundamental issue, grounded in a 
generative, hierarchical Bayesian framework.  The method advocated here constructs a sophisticated 
likelihood function, employing a non-trivial covariance matrix to treat the correlated 
pixel-to-pixel residuals generated from intrinsically imperfect models.  That matrix is composed of 
a linear combination of {\it global} (stationary) and {\it local} (non-stationary) Gaussian 
process kernels, which parameterize an overall mild covariance structure as well as small patches 
of highly discrepant outlier features.  The approach we describe is generally applicable to any 
spectroscopic inference problem (e.g., population synthesis in unresolved star clusters or 
galaxies, physical/chemical models of emission line spectra in star-forming regions, etc.).  
Moreover, it has the flexibility to incorporate additional information (as priors) or parametric 
complexity (if desired), and could be deployed as a substitute for a simplistic $\chi^2$ metric in 
already-established fitting tools (e.g., {\tt SME}).  To demonstrate how it is used, we determined 
the surface parameters of main-sequence stars with mid-F and mid-M spectral types from high-S/N 
optical and near-infrared data, with reference to pre-computed model libraries 
(Sect.~\ref{sec:examples}).  The source code developed here is open and freely available for use: 
see \url{http://github.com/iancze/Starfish}.

@@

\begin{figure*}[!t]
  \begin{center}
  \includegraphics{figs/pca_reconstruct.pdf}
  \figcaption{\emph{top} : The mean spectrum, standard deviation spectrum, and five eigenspectra that form the basis of the \textsc{Phoenix} synthetic library used to model Gl51. \emph{bottom}: The original synthetic spectrum from the \textsc{Phoenix} library compared with a spectrum reconstructed from a linear combination of the eigenspectra using Eqn~\ref{eqn:reconstruct} with the weights listed in the figure.
\label{fig:pca_reconstruct}}
  \end{center}
\end{figure*}

The first step in designing an emulator is to decompose the spectral library into an appropriate basis for emulation \citep{habib07, heitmann09}. We chose the principal component basis (PCA) to decompose the synthetic library into a set of ``eigenspectra,'' following the techniques of \citet{ivezic13}. Prior to the PCA decomposition, we first identify a subset of $M$ synthetic spectra which will be relevant to the star at hand. We then standardize the synthetic library by first subtracting off the mean spectrum and then ``whiten'' the spectra by dividing by the standard deviation measured in each pixel across the grid. The mean spectrum is 
\begin{equation}
  \xi_\mu = \frac{1}{N} \sum_{i = 1}^N f_\lambda(\{\vt_\ast \}^\textrm{grid})
\end{equation}
and the standard deviation spectrum is
\begin{equation}
  \xi_\sigma = \sqrt{\frac{1}{N} \sum_{i=1}^N \bigl [ f_\lambda(\{\vt_\ast \}^\textrm{grid}) - \xi_\mu \bigr]^2 } 
\end{equation}
All spectra in the library are then standardized by subtracting the mean spectrum and dividing by the standard deviation spectrum
\begin{equation}
  \widetilde{f}_\lambda(\{\vt_\ast \}^\textrm{grid}) = \frac{f_\lambda(\{\vt_\ast \}^\textrm{grid}) - \xi_\mu}{\xi_\sigma}
\end{equation}
Then, the eigenspectra are computed from the standardize grid using PCA analysis (\cite{ivezic13}). We decide to truncate our basis to the first $m$ of the eigenspectra, where $m$ is decided by the minimum number of eigenspectra required to reproduce any spectrum in the grid to better than 2\% accuracy for all pixels; however the mean pixel error is generally much smaller than this ($\lesssim 0.5\%$). We denote a single $N_{\rm pix}$ eigenspectrum as $\xi_k$, with principal component index $k = \{1, 2, \ldots, m\}$. As an example, the eigenspectra basis computed for Gl51 using the \textsc{Phoenix} library is shown in Fig~\ref{fig:pca_reconstruct}, top panel.

Using the PCA basis, we can lossily reconstruct any spectrum from the library with a linear combination of the eigenspectra
\begin{equation}
  f_\lambda(\{\vt_\ast \}^\textrm{grid}) \approx \xi_\mu + \xi_\sigma \sum_{k=1}^m w_k(\{\vt_\ast \}^\textrm{grid}) \xi_k
  \label{eqn:reconstruct}
\end{equation}
where $w_k$ is the weight of a given eigenspectrum, $k$. This weight is a three dimensional scalar function which depends on the stellar parameters $\vt_\ast$. We can determine the weights at the grid points in the library by taking the dot product of the standardized synthetic spectrum with each of the eigenspectra. The weight for a given eigenspectrum $\xi_k$ is 
\begin{equation}
  w_k(\{\vt_\ast \}^\textrm{grid}) = \sum_\lambda \widetilde{f}_\lambda(\{\vt_\ast \}^\textrm{grid}) \xi_k
\end{equation}
which is generally a smooth function of stellar parameters. To simplify the notation, we can write the weights in a length-$m$ column vector
\begin{equation}
  \mathbf{w} = 
  \begin{bmatrix}
    w_1\\
    w_2\\
    \vdots\\
    w_m
  \end{bmatrix}
\end{equation}
and horizontally concatenate the eigenspectra into a matrix with $N_{\rm pix}$ rows and $m$ columns
\begin{equation}
  \mathbf{\Xi} = 
  \begin{bmatrix}
    \xi_1 & \xi_2 & \cdots & \xi_m \\
  \end{bmatrix}
\end{equation}
Then we can rewrite Equation~\ref{eqn:reconstruct} as
\begin{equation}
  f_\lambda(\{\vt_\ast \}^\textrm{grid}) \approx \xi_\mu + \xi_\sigma I_{N_{\rm pix}} \mathbf{\Xi} \mathbf{w}
\end{equation}
Where $I_{N_{\rm pix}}$ is an $N_{\rm pix}$ by $N_{\rm pix}$ identity matrix. To recapitulate thus far, we have decomposed the spectra in the synthetic library into a principal component basis, allowing us to lossily reconstruct any spectrum in the synthetic library as a weighted linear combination of $m$ eigenspectra. The weight corresponding to each eigenspectrum is a moderately-smooth scalar function of the three stellar parameters. Therefore, to create a spectrum corresponding to a value of stellar parameters which is not in the spectral library, we must interpolate the weights to this new value. In practice, it might be possible to use a traditional interpolation scheme such as spline interpolation to interpolate the weights directly. However, we found that with high-signal to noise spectra (e.g., Gl51 is S/N $> 400$), the error in any given interpolated representation of the spectrum could constitute a significant portion of the uncertainty budget. This, combined with the Nyquist undersampling of the synthetic grid caused behavior where stellar parameters corresponding to a grid point in the synthetic library might be preferentially favored and bias the results, even after empirically accounting for interpolation error by doing ``drop-out'' interpolation tests. [This is more of a reason to use PCA vs. regular spectra, right?]

Therefore, we decide to develop a Bayesian spectral emulator (cite Habib, Heitmann) which allows us to propagate the uncertainty in interpolation forward to the likelihood calculation. First, we describe how we use a Gaussian process to model the interpolation of the weights then we marginalize over all possible interpolations to arrive at a modified likelihood function.
